 

### 2021-07-13 vertical-thermal-review.Rmd
working on ecology section

### Demography

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* 
Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors. 
Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a]. 
However, growth also declines as trees age. 
This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts]. 
Vertical biophysical conditions are bound to influence growth and whole tree functioning.

Greater $VPD$ and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes can cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

@@

Phenologically, in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].  However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more severly than in understory saplings that are buffered [@filewodImpactsSpringHeat2014]. Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].

>>>>

### Tree metabolism and growth

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeira_sizerelated_2015].
Specifically, increases are linked in part to increased leaf area and in part to increasing probability that the crown is in the canopy [@Muller-Landau_2006_EcologyLetters], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood:leaf area ratio (**REFS** -- [@IncreasingRespiratoryCosts]?).

- C allocation --> respiration, woody growth, NSCs, foliage turnover, belowground allocation, reproduction
- we don't know a whole lot about how C allocation scales with tree size
- woody growth increases with DBH to a point, then declines [@AT_inReview].
- NSCs increase , making smaller trees more sensitive to biological and biophyical stressors [@niinemetsResponsesForestTrees2010a]

**Vertical gradients also affect the climate sensitivity of metabolism and growth.**

- Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].

Greater $VPD$ and low RH can be more stressful for trees than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes can cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

@@

Phenologically, in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].  However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more severly than in understory saplings that are buffered [@filewodImpactsSpringHeat2014]. Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
